---
title: "varying pH analysis"
author: "Dan Stoebel"
edits: "Rodrigo Rosas"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
library(tidyverse)
library(readxl)
library(lubridate)
library(growthrates)
theme_set(theme_classic())
```

## Step 1: Import the data

```{r}
data_raw1 <- read_excel("2023-06-22 varying pH trial 1 (2).xlsx", range = "B28:CU317")
data_raw2 <- read_excel("2023-06-30 varying pH trial 2 (2).xlsx", range = "B28:CU221")
data_raw3 <- read_excel("2023-07-05 varying pH trial 3 (2).xlsx", range = "B28:CU221")
data_raw4 <- read_excel("2023-07-14 varying pH trial 4_1(success).xlsx", range = "B28:CU221")
```

## Step 2: Define key varibales

```{r}
blank_wells <- c("A1", "B1" , "C1", "D1", "E1", "F1", "G1", "H1", "A12", "B12", "C12", "D12", "E12", "F12","G12", "H12")

edge_cells <- c("A2", "A3", "A4", "A5", "A6", "A7", "A8", "A9", "A10", "A11","H2", "H3", "H4", "H5", "H6", "H7", "H8", "H9", "H10", "H11")

condition_1 <- c("A1", "A2", "A3", "A4", "A5", "A6","B1", "B2", "B3", "B4", "B5", "B6", "C1", "C2", "C3", "C4", "C5", "C6", "D1", "D2", "D3", "D4", "D5", "D6", "E1", "E2", "E3", "E4", "E5", "E6", "F1", "F2", "F3", "F4", "F5", "F6", "G1", "G2", "G3", "G4", "G5", "G6", "H1", "H2", "H3", "H4", "H5", "H6")
```

## Wrangle the data into 'tidy format'

```{r}
data_finaltrial1 <- data_raw1 %>% 
  pivot_longer(cols = matches("[A-H][0-9]+"), names_to = "well", values_to = "OD600") %>% 
  group_by(well) %>% 
  mutate(elapsed_mins = (Time[[1]] %--% Time) / dminutes(1) ) %>% 
  ungroup() %>%
  mutate(well_status = case_when(well %in% blank_wells ~ "blank",
                                 well %in% edge_cells ~ "edge",
                                 TRUE ~ "interior")) %>%
  mutate(condition = case_when(well %in% condition_1 ~ "initial of 6.50", TRUE ~ "initial of 5.99")) %>%
  select(elapsed_mins, well, well_status, OD600, condition)
  
data_final2trial1 <- data_finaltrial1 %>%
  filter(elapsed_mins <= 1920) %>%
  group_by(elapsed_mins, condition) %>%
  summarise(avg_od = mean(OD600))
```
```{r}
data_finaltrial2 <- data_raw2 %>% 
  pivot_longer(cols = matches("[A-H][0-9]+"), names_to = "well", values_to = "OD600") %>% 
  group_by(well) %>% 
  mutate(elapsed_mins = (Time[[1]] %--% Time) / dminutes(1) ) %>% 
  ungroup() %>%
  mutate(well_status = case_when(well %in% blank_wells ~ "blank",
                                 well %in% edge_cells ~ "edge",
                                 TRUE ~ "interior")) %>%
  mutate(condition = case_when(well %in% condition_1 ~ "initial of 5.50", TRUE ~ "initial of 5.05")) %>%
  select(elapsed_mins, well, well_status, OD600, condition)
  
data_final2trial2 <- data_finaltrial2 %>%
  group_by(elapsed_mins, condition) %>%
  summarise(avg_od = mean(OD600))
```

```{r}
data_finaltrial3 <- data_raw3 %>% 
  pivot_longer(cols = matches("[A-H][0-9]+"), names_to = "well", values_to = "OD600") %>% 
  group_by(well) %>% 
  mutate(elapsed_mins = (Time[[1]] %--% Time) / dminutes(1) ) %>% 
  ungroup() %>%
  mutate(well_status = case_when(well %in% blank_wells ~ "blank",
                                 well %in% edge_cells ~ "edge",
                                 TRUE ~ "interior")) %>%
  mutate(condition = case_when(well %in% condition_1 ~ "initial of 4.48", TRUE ~ "initial of 4.07")) %>%
  select(elapsed_mins, well, well_status, OD600, condition)
  
data_final2trial3 <- data_finaltrial3 %>%
#  filter(elapsed_mins >= 100) %>%
  group_by(elapsed_mins, condition) %>%
  summarise(avg_od = mean(OD600))
```

```{r}
data_finaltrial4 <- data_raw4 %>% 
  pivot_longer(cols = matches("[A-H][0-9]+"), names_to = "well", values_to = "OD600") %>% 
  group_by(well) %>% 
  mutate(elapsed_mins = (Time[[1]] %--% Time) / dminutes(1) ) %>% 
  ungroup() %>%
  mutate(well_status = case_when(well %in% blank_wells ~ "blank",
                                 well %in% edge_cells ~ "edge",
                                 TRUE ~ "interior")) %>%
  mutate(condition = case_when(well %in% condition_1 ~ "initial of 7.04", TRUE ~ "initial of 7.54")) %>%
  select(elapsed_mins, well, well_status, OD600, condition)
  
data_final2trial4 <- data_finaltrial4 %>%
  group_by(elapsed_mins, condition) %>%
  summarise(avg_od = mean(OD600))
```
## Join all the indivdual tidy data frames to one tibble

```{r}
allthedata <- bind_rows(data_final2trial1, data_final2trial2, data_final2trial3, data_final2trial4)
```

##make 'condtion' a factor and reorder for increasing pH on legend

```{r}
allthedata2 <- allthedata
allthedata2$condition <- as.factor(allthedata2$condition)
allthedata2$condition <-factor(allthedata2$condition, levels = c("initial of 4.07", "initial of 4.48", "initial of 5.05", "initial of 5.50", "initial of 5.99", "initial of 6.50", "initial of 7.04", "initial of 7.54"))
```

```{r}
allthedata2 %>%
  ggplot(aes(x = elapsed_mins, y = avg_od)) + 
  geom_point(aes(color= condition)) +
  scale_color_manual(values=c("#FD4027","#FDB727","#FFF833","#3EFF33","#196F3D", "#27FDED", "#0256D4", "#602DC1"))+
  theme(legend.text=element_text(size=12))
```
# Analysis of growth rates

## Initial visuzlization

Before we try to estimate a growth rate, let's just take a look at our data and make sure things look sensible.

```{r}
data_final %>% 
  ggplot(aes(x = elapsed_mins, y = OD600)) +
  geom_point(alpha = 0.1) +
  facet_wrap(condition ~ well_status)
```
That looks good! We see none of the blanks grew, and all of the wells with cells look like they grew. It looks like things might be slightly more variable on the edge than in the interior.


The package `growthrates` has many ways to estimate growth rates. Type `vignette("Introduction", package = "growthrates")` on the console to learn lots more.

For this analysis we'll use the "easy linear" approach, which uses liner regression to find the maximal growth rate.
```{r}
splitted.data <- multisplit(data_final2, "condition")
data_finalcond1 <- splitted.data[[1]]
data_finalcond2 <- splitted.data[[2]]
```

```{r}
cells_only <- filter(data_final, well_status != "blank")

many_easylinear_fits <- all_easylinear(OD600 ~ elapsed_mins |well , data = cells_only)
```

```{r}
fit_cond1 <- fit_easylinear(data_finalcond1$elapsed_mins, data_finalcond1$avg_od)
fit_cond2 <- fit_easylinear(data_finalcond2$elapsed_mins, data_finalcond2$avg_od)
yint_cond1 <- coef(fit_cond1)[[1]]
slope_cond1 <- coef(fit_cond1)[[3]]
yint_cond2 <- coef(fit_cond2)[[1]]
slope_cond2 <- coef(fit_cond2)[[3]]
```

```{r}
data_final2 %>%
  ggplot(aes(x = elapsed_mins, y = avg_od)) + 
  geom_point(aes(color= condition)) +
  geom_abline(slope = slope_cond1, intercept = yint_cond1) +
  geom_abline(slope = slope_cond2, intercept = yint_cond2)
```

Plot out each of these individual cases.
```{r}
#pdf(width = 30, height = 30)
par(mfrow = c(1,2))
par(mar = c(2.5, 4, 2, 1))
plot(fit_cond1, log = "y")
plot(fit_cond1)
#dev.off()
```


Calculate some summary statistics
```{r}
growth_rates <- coef(many_easylinear_fits) %>% 
  as_tibble(rownames = "well") %>% 
    mutate(well_status = case_when(well %in% blank_wells ~ "blank",
                                 well %in% edge_cells ~ "edge",
                                 TRUE ~ "interior")) 

growth_rates %>% 
  group_by(well_status) %>% 
  summarise(mean_mumax = mean(mumax),
            sd_mumax = sd(mumax))
```


```{r}
growth_rates %>% 
  mutate(condition = case_when(well %in% condition_1 ~ "inital of 5.50", TRUE ~ "intial of 5.05")) %>%
  ggplot(aes(x = well_status, y = mumax)) +
  geom_boxplot(outlier.shape = NA, color = "gray") +
  geom_jitter(width = .1) +
  facet_wrap(~condition)
```
There is clearly a difference between the edges and the interiors of the plate. I'll construct a confidence interval to get a sense of how big the difference is.

```{r}
library(infer)

diff_growth_rates <- growth_rates %>% 
  specify(mumax ~ well_status) %>% 
  calculate(stat = "diff in means", order = c("interior", "edge"))

growth_rates %>% 
  specify(mumax ~ well_status) %>% 
  generate(reps = 1000, type = "bootstrap") %>% 
  calculate(stat = "diff in means", order = c("interior", "edge")) %>% 
  get_ci()

```

