---
title: "varying pH analysis"
author: "Dan Stoebel"
edits: "Rodrigo Rosas"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
library(tidyverse)
library(readxl)
library(lubridate)
library(growthrates)
theme_set(theme_classic())
```

## Step 1: Import the data

The data are in an excel file. The output is not in a rectangular format, which is not ideal for R. However, there is one rectangular piece of the data that has all we want for our analysis - cells B28 to CU185. I can read in just this chunk of data with:
```{r}
data_raw <- read_excel("2023-07-05 varying pH trial 3 (2).xlsx", range = "B28:CU221")
```


```{r}
blank_wells <- c("A1", "B1" , "C1", "D1", "E1", "F1", "G1", "H1", "A12", "B12", "C12", "D12", "E12", "F12","G12", "H12")

edge_cells <- c("A2", "A3", "A4", "A5", "A6", "A7", "A8", "A9", "A10", "A11","H2", "H3", "H4", "H5", "H6", "H7", "H8", "H9", "H10", "H11")

condition_1 <- c("A1", "A2", "A3", "A4", "A5", "A6","B1", "B2", "B3", "B4", "B5", "B6", "C1", "C2", "C3", "C4", "C5", "C6", "D1", "D2", "D3", "D4", "D5", "D6", "E1", "E2", "E3", "E4", "E5", "E6", "F1", "F2", "F3", "F4", "F5", "F6", "G1", "G2", "G3", "G4", "G5", "G6", "H1", "H2", "H3", "H4", "H5", "H6")
```

```{r}
data_final <- data_raw %>% 
  pivot_longer(cols = matches("[A-H][0-9]+"), names_to = "well", values_to = "OD600") %>% 
  group_by(well) %>% 
  mutate(elapsed_mins = (Time[[1]] %--% Time) / dminutes(1) ) %>% 
  ungroup() %>%
  mutate(well_status = case_when(well %in% blank_wells ~ "blank",
                                 well %in% edge_cells ~ "edge",
                                 TRUE ~ "interior")) %>%
  mutate(condition = case_when(well %in% condition_1 ~ "inital of 4.5", TRUE ~ "intial of 4.0")) %>%
  select(elapsed_mins, well, well_status, OD600, condition)
  
```

```{r}
data_final2 <- data_final %>%
  filter(well_status == "interior") %>%
  filter(elapsed_mins >= 100) %>%
  group_by(elapsed_mins, condition) %>%
  summarise(avg_od = mean(OD600))
```
```{r}
data_final2 %>%
  ggplot(aes(x = elapsed_mins, y = avg_od)) + 
  geom_point(aes(color= condition))
```
# Analysis of growth rates

## Initial visuzlization

Before we try to estimate a growth rate, let's just take a look at our data and make sure things look sensible.

```{r}
data_final %>% 
  ggplot(aes(x = elapsed_mins, y = OD600)) +
  geom_point(alpha = 0.1) +
  facet_wrap(condition ~ well_status)
```
That looks good! We see none of the blanks grew, and all of the wells with cells look like they grew. It looks like things might be slightly more variable on the edge than in the interior.


The package `growthrates` has many ways to estimate growth rates. Type `vignette("Introduction", package = "growthrates")` on the console to learn lots more.

For this analysis we'll use the "easy linear" approach, which uses liner regression to find the maximal growth rate.
```{r}
splitted.data <- multisplit(data_final2, "condition")
data_finalcond1 <- splitted.data[[1]]
data_finalcond2 <- splitted.data[[2]]
```

```{r}
cells_only <- filter(data_final, well_status != "blank")

many_easylinear_fits <- all_easylinear(OD600 ~ elapsed_mins |well , data = cells_only)
```

```{r}
fit_cond1 <- fit_easylinear(data_finalcond1$elapsed_mins, data_finalcond1$avg_od)
fit_cond2 <- fit_easylinear(data_finalcond2$elapsed_mins, data_finalcond2$avg_od)
yint_cond1 <- coef(fit_cond1)[[1]]
slope_cond1 <- coef(fit_cond1)[[3]]
yint_cond2 <- coef(fit_cond2)[[1]]
slope_cond2 <- coef(fit_cond2)[[3]]
```

```{r}
data_final2 %>%
  ggplot(aes(x = elapsed_mins, y = avg_od)) + 
  geom_point(aes(color= condition)) +
  geom_abline(slope = slope_cond1, intercept = yint_cond1) +
  geom_abline(slope = slope_cond2, intercept = yint_cond2)
```

Plot out each of these individual cases.
```{r}
#pdf(width = 30, height = 30)
par(mfrow = c(1,2))
par(mar = c(2.5, 4, 2, 1))
plot(fit_cond1, log = "y")
plot(fit_cond1)
#dev.off()
```


Calculate some summary statistics
```{r}
growth_rates <- coef(many_easylinear_fits) %>% 
  as_tibble(rownames = "well") %>% 
    mutate(well_status = case_when(well %in% blank_wells ~ "blank",
                                 well %in% edge_cells ~ "edge",
                                 TRUE ~ "interior")) 

growth_rates %>% 
  group_by(well_status) %>% 
  summarise(mean_mumax = mean(mumax),
            sd_mumax = sd(mumax))
```


```{r}
growth_rates %>% 
  ggplot(aes(x = well_status, y = mumax)) +
  geom_boxplot(outlier.shape = NA, color = "gray") +
  geom_jitter(width = .1)
```
There is clearly a difference between the edges and the interiors of the plate. I'll construct a confidence interval to get a sense of how big the difference is.

```{r}
library(infer)

diff_growth_rates <- growth_rates %>% 
  specify(mumax ~ well_status) %>% 
  calculate(stat = "diff in means", order = c("interior", "edge"))

growth_rates %>% 
  specify(mumax ~ well_status) %>% 
  generate(reps = 1000, type = "bootstrap") %>% 
  calculate(stat = "diff in means", order = c("interior", "edge")) %>% 
  get_ci()

```

